Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-11 rnni_convexity.tex
move NP-hardness things from Split Theorem to Discussion
rm commeted out tex with some ideas and importantly LATTICES -- @lenacoll please check you don't need these and add the lattices section back
discussion at the beginning of section:cluster_theorem

\todo{Why are we in the Geometry section then?}
In this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.

The proof of $\np$-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.

\begin{theorem}
There are trees $T,R$ in $\nni$ sharing a cluster which is not shared by any intermediate tree on any shortest path from $T$ to $R$.
\label{thm:split_nni}
\end{theorem}

\begin{proof}
See \autocite{Li1996-zw}.
\end{proof}

In the proof of Theorem~\ref{thm:split_nni} presented in \autocite{Li1996-zw} an example of two trees $T$ and $R$ that share clusters $C$ and $X \setminus C$ but have no shortest path that preserves these cluster is given.
A simplified illustration of this example is provided in Figure~\ref{fig:NNI_NP_proof}.
\autocite{Li1996-zw} prove that on a shortest path between $T$ and $R$ in $\nni$ at first $T$ is transformed to a tree $T'$ with $\frac{n}{2}$ cherries, each of which contains one taxon of each of the clusters $C$ and $X \setminus C$.
Then $T'$ is sorted to a tree $R'$ containing the same cherries as $T'$ such that that resolving all these cherries of $R'$ results in $R$.
Specifically, \autocite{Li1996-zw} prove that there is no shortest path between $T$ and $R$ in $\nni$ where $C$ and $X \setminus C$ are present in all trees on that path.

The reason that this example cannot be transferred to $\rnni$ is that there are lots of additional rank moves needed on a path between $T$ ans $R$ similar to the one described above.
As in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is longer than a shortest $\rnni$ path.
Sorting the two caterpillar subtrees of $T$ and $R$ separately results in a shorter path in $\rnni$.

@@

% \todo{LC: We still need to check 7 taxa.}
% We are doing this in the following way.
% We focus on the subset of trees that induce share a cluster.
% It is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$, because of the symmetry of the space:.
% distances $d(T_1,R_1)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
% Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all pairs of trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra's algorithm \autocite{Dijkstra1959-ph} on the $\rnni$ graph, which we can compute as explained in Section~\ref{section:alg_RNNI_graph}.
% Additionally, we compute the subgraph of $\rnni$ induced by the subset of trees that include the cluster $\{1, \ldots, m\}$.
% With the Floyd-Warshall algorithm we compute all pairwise distances between the trees in this subgraph and compare them with the distances in $\rnni$.
% With this approach it is possible to show that the Cluster Theorem holds for up to seven taxa.
% \todo{LC: We still need to check 7 taxa.}

% \section{Ideas}
%
% [This section includes some ideas that we could possibly include in the paper]
%
% Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
%
% \begin{proposition}
% If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
% \label{proposition:lower_bound_distance}
% \end{proposition}
%
%
% The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
% $\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
% An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
% However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
%
% \begin{conjecture}
% Let $T$ and $R$ be ranked trees.
% The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
% \end{conjecture}
%
% \begin{figure}[H]
% \centering
% \includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
% \vspace{12pt}
% \caption{Two paths computed by $\findpath$.
% At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
% \label{fig:findpath_not_symmetric}
% \end{figure}
%
% % idea for description of max distance caterpillar trees:
% \begin{lemma}
% Let $T$ and $R$ be caterpillar trees.
% It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
% \todo{define induced triplets}
% \end{lemma}
%
% This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
% Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
%
% \begin{proof}
% %TODO
% % Insert the proof here
% Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
% \end{proof}
%
% According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
% The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
%
%
% \subsection{Partition lattice}
%
% % Define Partition Lattice $\Pi_n$ and max chains in that lattice!
%
% \begin{theorem}
% The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
% The corresponding metric spaces are isometric.
% \end{theorem}

@@

The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.
Before investigating geometrical properties of $\rnni$ we introduced some algorithms in Section~\ref{section:algorithms}).
One of the most important algorithms we introduced here is $\findpath$, which can be used for approximating $\rnni$ distances.
We could in particular show that this algorithm computes exact distances for all pairs of trees on up to six taxa.
This algorithm might be a useful tool for future investigations of the $\rnni$ space.

>>>>

In this section we discussed the so-called Split Theorem, which states that every split shared between two trees is maintained along shortest paths between the trees.
For example, Split Theorem is true in the $\spr$ space and false in $\nni$ \autocite{Li1996-zw}.
Specifically, in $\nni$ there exist two trees $T$ and $R$, which share a split $A | B$, such that every tree on every shortest path between $T$ and $R$ does not have $A | B$.

To construct this example, \textcite{Li1996-zw} used the idea illustrated in Figure~\ref{fig:NNI_NP_proof} and showed that for an appropriate permutation of the taxa in the two caterpillar subtrees of both $T$ and $R$, it is strictly shorter to merge the two caterpillar subtrees first, then sort them simultaneously, and then split back to two caterpillar subtrees, rather the two caterpillar subtrees independently.

The reason that this example cannot be transferred to $\rnni$ is that additional rank moves are necessary to merge and then split the two caterpillar subtrees.
Similarly to the example in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is no longer a shortest $\rnni$ path.
Sorting the two caterpillar subtrees of $T$ and $R$ independently results in a shorter path in $\rnni$.

@@

@@

Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.

The proof of $\np$-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.
Before investigating geometrical properties of $\rnni$ we introduced some algorithms in Section~\ref{section:algorithms}).
One of the most important algorithms we introduced here is $\findpath$, which can be used for approximating $\rnni$ distances.
We could in particular show that this algorithm computes exact distances for all pairs of trees on up to six taxa.
This algorithm might be a useful tool for future investigations of the $\rnni$ space.
